# Executive Summary
Through 2010, octopuses were managed as part of the Bering Sea/Aleutian Islands (BSAI) "other species" complex, along with sharks, skates, and sculpins. Historically, catches of the other species complexes were well below TAC. Due to increasing market values, retention of species within the other species complex increased. Beginning in 2011, an amendment to the BSAI fisheries management plan provided separate management for sharks, skates, sculpins, and octopus and set catch limits for each species group. Initially, catch limits for octopus were set using Tier 6 methods based on the maximum historical incidental catch rate. Since 2012, a methodology based on consumption of octopus by Pacific cod (*Gadus macrocephalus*) has been used to set catch limits  (see @conners2016assessment for methodological details). 

This year’s assessment is an operational update, meaning new consumption data was provided through 2023 to determine catch limits using the previous assessment's model/methodology. Alternative models/methodologies were not considered. At least eight species of octopus are found in the BSAI though in this update assessment, all octopus species are grouped into a single assemblage. The species composition of the octopus community is not well documented, but data indicate that the giant Pacific octopus (*Enteroctopus dofleini*) is the most common (Tables `r params$chapter`-\@ref(tab:tab-AI-survey),-\@ref(tab:tab-shelf-survey),-\@ref(tab:tab-slope-survey)). Octopuses are taken as incidental catch in trawl, longline, and pot fisheries with a portion retained and sold for human consumption or bait. The BSAI trawl surveys produce highly variable biomass estimates for octopus (Figure `r params$chapter`-\@ref(fig:fig-survey-bio)). 



## Summary of Changes in Assessment Inputs

*Changes in Input Data*

1. The calculation of annual and long-term average consumption rates has been updated using 13,614 additional Pacific cod stomach samples collected from 2012-2013 and 2016-2023.

*Changes in Assessment Methodology* \
There have been no changes to the assessment methodology


## Summary of Results



For `r year + 1`, the recommended maximum allowable ABC from the Tier 6 specifications is `r format(round(mABC_vec[3]), big.mark=",")` t. Reference values for octopus are summarized in the following table, with the recommended ABC and OFL values for `r year + 1` in bold.

```{r, child= "01a-main_table.Rmd", eval = T}

```

The BSAI octopus complex is not currently subject to overfishing because the `r year-1` catch, the most recent completed year, of `r format(round(prior_cat), big.mark=",")` t is smaller than the `r year-1` OFL of `r format(round(prior_OFL), big.mark=",")` t. 


```{r, child= "01c-plan_team.Rmd", eval = F}

```
