## Risk Table and ABC recommendation

The following template is used to complete the risk table:
```{r}
library(flextable)
library(tidyverse)


data.frame(level = c("Level 1: No Conern",
                     "Level 2: Major Concern",
                     "Level 3: Extreme concern"),
           assess = c("Typical to moderately increased uncertainty/minor unresolved issues in assessment.", "Major problems with the stock assessment; very poor fits to data; high level of uncertainty; strong retrospective bias.",
                      "Severe problems with the stock assessment; severe retrospective bias. Assessment considered unreliable."),
           pop = c("Stock trends are typical for the stock; recent recruitment is within normal range.",
                   "Stock trends are highly unusual; very rapid changes in stock abundance, or highly atypical recruitment patterns.",
                   "Stock trends are unprecedented; More rapid changes in stock abundance than have ever been seen previously, or a very long stretch of poor recruitment compared to previous patterns."),
           enviro = c("No apparent environmental/ecosystem concerns",
                      "Multiple indicators showing consistent adverse signals a) across the same trophic level as the stock, and/or b) up or down trophic levels (i.e., predators and prey of the stock)",
                      "Extreme anomalies in multiple ecosystem indicators that are highly likely to impact the stock; Potential for cascading effects on other ecosystem components"),
           fish = c("No apparent fishery/resource-use performance and/or behavior concerns",
                    "Multiple indicators showing consistent adverse signals a) across different sectors, and/or b) different gear types",
                    "Extreme anomalies in multiple performance  indicators that are highly likely to impact the stock")) %>% 
  flextable() %>% 
    font(fontname = "Times", part = "all") %>% 
    width(j = ~ assess, width = 1.5) %>% 
    width(j = ~ pop, width = 1.5) %>% 
    width(j = ~ enviro, width = 1.5) %>% 
    width(j = ~ fish, width = 1.5) %>% 
    set_header_labels(level = "",
                    assess = "Assessment-related considerations",
                    pop = "Population dynamics considerations",
                    enviro = "Environmental/ecosystem considerations",
                    fish = "Fishery Performance") %>% 
  font(fontname = "Times", part = "all") %>% 
  fontsize(size = 10) %>% 
  theme_vanilla() %>% 
  italic(part = "header") 
```  
The table is applied by evaluating the severity of four types of considerations that could be used to support a scientific recommendation to reduce the ABC from the maximum permissible. 
These considerations are stock assessment considerations, population dynamics considerations, environmental/ecosystem considerations, and fishery performance.

*Assessment-related considerations*: The octopus complex is a Tier 6 stock and thus by definition data-limited; many potential concerns such as the lack of reliable abundance data are inherent to Tier 6 stocks and harvest recommendations are made accordingly. There are no additional considerations that would warrant reducing the ABC below maximum permissible. Rated Level 1, normal.

*Population dynamics considerations*: For a data-limited stock such considerations are hard to evaluate. In addition octopuses are short-lived and signals like recruitment are harder to discern. However, the results from this years operational updated have an increased OFL and ABC from the previous year. Rated Level 1, normal.

*Environmental/ecosystem considerations*: Very little is understood about mechanistic linkages between ecosystem drivers and octopus growth, survival, or stock dynamics. Octopus are fast-growing animals. @Sano2018 studied the giant Pacific octopus (*Enteroctopus dolfeini*) in captivity and found both growth and feeding rates have a dome shaped relationship with temperature where maximum growth rates have been observed around 10°C and maximum feeding rates between 10-15°C; activity also tends to increase with temperature. Limited knowledge of these species exists to identify stock-specific indicators. Peak spawning occurs in winter to early spring, hatch time varies between 150 days to one year and the paralarvae period of ~90 days occurs around June- Aug [@Conrath2014]. Based on 1993-2022 bycatch records (North Pacific Observers groundfish data, https://apps-afsc.fisheries.noaa.gov/ords/r/fma_ols/fma-map/) and frequency of occurrence in non-empty Pacific cod stomachs from 1981-2011 [@Rohan2017] show octopus are distributed along the outer and middle domain, and the Aleutian Islands. @Spencer2019 assessed the climate vulnerability of Pacific giant octopus based on traits and expert evaluation, and found the sensitivity, exposure and vulnerability to be low when rubrics were scored moderate only if 2 or more individual attributes had mean values ≥ 2.5. However, when an alternative logic rule was used (less than 2 attributes with mean value ≥ 2.5 and more than ≥2 attributes with mean value ≥ 1.5), then the ranking increased to moderate for sensitivity, exposure and vulnerability.


[**Environmental processes**]{.underline}: 
In 2023, broad-scale climate patterns, like the North Pacific Index, reflected a transition from La Niña conditions to developing El Niño conditions in the tropic Pacific. In the eastern Bering Sea, regional sea surface temperature trends were at or near the long-term average in 2023. Exceptions to near-normal sea surface temperature (SST) conditions include a relatively warm winter across the shelf. Above-average SSTs lasted through spring over the outer (100-200m isobaths) and middle (50-100m isobaths) domains. Bottom temperatures derived from the ROMS model showed consistently cooler than average bottom temperatures over the outer domain (100-200m) from September 2022 through August 2023. Sea ice metrics, such as early ice extent (Oct. - Dec.), annual ice extent, and sea ice thickness were all near the respective time series averages. The 2023 cold pool extent was also near its historical average [@Hennon2023]. Data from the 2023 EBS bottom trawl survey (AFSC 2023) show bottom temperatures below 6°C in the outer and middle domain. Likewise, bottom temperatures in the AI from 1994-2022 show mean bottom temperatures across the chain below 6°C. Temperatures from the longline survey show temperature between 100 - 300 m was also below 6°C in the eastern Aleutians [@Siwicke2023]. The trend of increasing temperatures in the AI favors faster growth rates and increased feeding rates. Within the EBS shelf, overall octopus biomass seems to be inversely related to cold pool areas, however it is unclear whether the changes in biomass are driven by population increase or movement of octopus (e.g. from the slope to the shelf). 

[**Prey**]{.underline}: 
Prey for octopus includes shrimps, crabs, and benthic infauna (e.g., clams). No direct measures for shrimps are available from the eastern Bering Sea or Aleutian Islands. In the eastern Bering Sea, biomass trends for several crab stocks have declined in recent years and are currently below their respective time series averages [@Richar2023]. Indirect measures of benthic infauna availability based on biomass trends of benthic foragers indicates reduced infaunal prey available [@Siddon2023]. Available information on prey resources for octopus is difficult to directly assess; some indicators of prey conditions appear low in 2023, though mechanistic linkages to octopus stock dynamics remain elusive. There is no prey information for the AI at this time.

[**Competitors**]{.underline}: 
Competitors of adult or large octopus include Pacific halibut and Pacific cod among others. The Pacific halibut stock decreased from a peak in the early 2000s and remains low in 2023, therefore represents no increase in competitor pressure [@Stewart2022]. In the EBS Pacific cod had a modest increase in biomass from 2022 to 2023, but also does not represent a substantial increase in competitor pressure. In the AI, total biomass of Pacific cod has been steadily decreasing with biomass in 2022 at ~25% of its peak biomass in 1989 [@Spies2022_assessment_pcod].

[**Predators**]{.underline}: 
Predators of octopus include Pacific cod, fur seals, Steller sea lions, and seabirds. Pacific cod had a modest increase in biomass from 2022 to 2023, while fish condition (based on length-weight residuals) over the outer shelf was below average in 2023 [@Prohaska2023]. Fur seal population trends at St. Paul Island continued a declining trend through 2022 [@Siddon2023]. Seabird populations at the Pribilof Islands showed mixed trends for 2023 with higher reproductive success at St. George compared to St. Paul Island [@Siddon2023].  Additional occasional predators of octopus in the Aleutian Islands include Pacific halibut and Atka mackerel as well as Pacific cod and Steller sea lions [@Ormseth2020_assessment_octopus; @Fritz2019).  Steller sea lion populations in the western and central Aleutian Islands continued a declining trend [@Sweeney2022] and Atka mackerel decreased slightly in 2022 to around 561,000 tons and is down from almost peak abundances in 2003-04 of 1 million tons. Overall predation on octopus in the Aleutians is expected to be lower than in past years.


*Summary for Environmental/Ecosystem considerations*: 

* **Environment**: Oceanographic metrics for the EBS showed average conditions over the past year (August 2022 - August 2023). In the Aleutian Islands midwater temperatures remained below 6°C (summer 2023) in the Eastern AI, as have mean bottom summer temperatures across the chain. 
* **Prey**: In the EBS, some indicators of prey conditions appear low in 2023, though mechanistic linkages to octopus stock dynamics remain elusive. No information in the AI.
*	**Competitors**: Trends in available information for competitors of octopus suggest no increase in competitor pressure in 2023 in the EBS and decreasing trends in the AI (of known competitors)
*	**Predators**: Trends in available information for predators of octopus suggest no increase in predation pressure in 2023 for the EBS and decreasing trends in the AI

Proper evaluation of risk is difficult for a data-limited stock. However, the available data suggest there are no apparent ecosystem concerns--Level 1.

  
  
*Fishery performance*: As a nontarget stock, catches of octopuses in the GOA are influenced by their abundance and by the behavior of target fisheries. The catches have remained low, with the majority of the catch occurring as bycatch in the Pacific cod fishery. Rated Level 1, normal.

#### Summary and ABC recommendation

```{r}
l1 = "Level 1: No increased concerns"
l2 = "Level 2: Substantially increased concerns"
l3 = "Level 3: Major concern"
l4 = "Level 4: Extreme concern"


data.frame(assess = l1,
           pop = l1,
           enviro = l1,
           fish = l1) %>% 
  flextable() %>% 
    width(j = ~ assess, width = 1.5) %>% 
    width(j = ~ pop, width = 1.5) %>% 
    width(j = ~ enviro, width = 1.5) %>% 
    width(j = ~ fish, width = 1.5) %>% 
    set_header_labels(level = "",
                    assess = "Assessment-related considerations",
                    pop = "Population dynamics considerations",
                    enviro = "Environmental/ecosystem considerations",
                    fish = "Fishery Performance") %>% 
  fontsize(size = 10) %>% 
  font(fontname = "Times", part = "all") %>% 
  italic(part = "header") 
```

The score of level 1 for each category suggests that setting the ABC below the maximum permissible is not warranted at this time.
